---
title: "R Notebook"
output: html_notebook
---

```{r}
library (tidyverse)
library (readxl)
library (modeldata)
library(dplyr)
library(corrplot)


#Importing Dataset
wine_data<-read_xlsx("C:/Users/kld00004/Downloads/Wine Quality Dataset.xlsx")


#Exploring Data
str(wine_data)
head(wine_data)
summary(wine_data)



#Setting Numeric and Factor Columns
wine_numeric_columns <- wine_data[, sapply(wine_data, is.numeric) & names(wine_data) != "type"]
wine_factor_columns <- sapply(wine_data$type, as.factor)
wine_data$type <- factor(wine_data$type, levels = c('white', 'red'))


# Setting up to filter data for red and white wines separately
red_wine <- wine_data %>% filter(type == "red")
white_wine <- wine_data %>% filter(type == "white")

#Count in each rating
count_red_quality <- red_wine %>% count(quality)
count_red_quality

count_white_quality <- white_wine %>% count(quality)
count_white_quality

```

```{r #boxplots}
#Checking for Outliers
#Combined Box plots
par(mfrow = c(2, 7), mfcol = c(2, 7))  # 2 rows, 7 columns

for (col in names(wine_data)) {
  boxplot(wine_data[[col]], main = col, col = "skyblue")}

#Red Wine Box plots
par(mfrow = c(2, 7), mfcol = c(2, 7))  # 2 rows, 7 columns

for (col in names(red_wine)) {
  boxplot(red_wine[[col]], main = col, col = "skyblue")}


#White Wine Box plots
par(mfrow = c(2, 7), mfcol = c(2, 7))  # 2 rows, 7 columns

for (col in names(white_wine)) {
  boxplot(white_wine[[col]], main = col, col = "skyblue")}

```




```{r}
##Cleaning Data of Outliers
# Identify and handle outliers for red wine
filtered_red_wine <- red_wine %>%
  select(-type)%>%
  filter_all(all_vars(between(., quantile(., 0.25) - 1.5 * IQR(.), quantile(., 0.75) + 1.5 * IQR(.))))

# Identify and handle outliers for white wine
filtered_white_wine <- white_wine %>%
   select(-type)%>%
  filter_all(all_vars(between(., quantile(., 0.25) - 1.5 * IQR(.), quantile(., 0.75) + 1.5 * IQR(.))))

# Combine the filtered red and white wines
filtered_wine_data <- bind_rows(filtered_red_wine, filtered_white_wine)


```

```{r}
# Filtered Box Plots

#Combined Filtered Box Plots
par(mfrow = c(2, 7), mfcol = c(2, 7))  # 2 rows, 7 columns

for (col in names(filtered_wine_data)) {
  boxplot(filtered_wine_data[[col]], main = col, col = "skyblue")
}

#Red Wine Filtered Box Plots
par(mfrow = c(2, 7), mfcol = c(2, 7))  # 2 rows, 7 columns

for (col in names(filtered_red_wine)) {
  boxplot(filtered_red_wine[[col]], main = col, col = "skyblue")
}


#White Wine Filtered Box Plots
par(mfrow = c(2, 7), mfcol = c(2, 7))  # 2 rows, 7 columns

for (col in names(filtered_white_wine)) {
  boxplot(filtered_white_wine[[col]], main = col, col = "skyblue")
}
```

```{r}
#Count in each rating - outliers removed
count_red_filtered_qual <- filtered_red_wine %>% count(quality)
count_red_filtered_qual


count_white_filtered_qual <- filtered_white_wine %>% count(quality)
count_white_filtered_qual

```

The majority of both types of wine fall between quality ratings 4 and 5.  with the lowest rating being 4 and highest 7. The outliers made up the entirety of the 3, 8, and 9 ratings.

Perhaps should analyze the difference between the filtered vs the low and high outliers.  There appears to be something in the outliers that make the quality excellent or poor.



## Analysis for the filtered data (Ratings 4-7)


```{r}

# Compute correlation matrices for red and white wines
cor_matrix_red <- cor(filtered_red_wine[, c(1:12)])
cor_matrix_white <- cor(filtered_white_wine[, c(1:12)])
cor_matrix_combined <-cor(filtered_wine_data)

# Cor Matrix Red
print(cor_matrix_red)

#Cor Matrix White
print(cor_matrix_white)

#Cor Matrix Combined
print(cor_matrix_combined)

# Plotting the correlation matrices side by side
par(mfrow = c(1, 2))

# Plot for red wine
corrplot(cor_matrix_red, method = "color", title = "Correlation Plot - Red Wine", mar=c(0,0,5,0))

# Plot for white wine
corrplot(cor_matrix_white, method = "color", title = "Correlation Plot - White Wine", mar=c(0,0,5,0))

# Plot for white wine
corrplot(cor_matrix_combined, method = "color", title = "Correlation Plot - Red & White Wine", mar=c(0,0,5,0))


```



```{r #Regressions}

red_model <-lm(filtered_red_wine$quality~ ., data = filtered_red_wine)
summary(red_model)

white_model <- lm(filtered_white_wine$quality ~ ., data = filtered_white_wine)
summary(white_model)

combined_model <- lm(filtered_wine_data$quality ~ ., data = filtered_wine_data)
summary(combined_model)


```


```{r #Stepped Regressions}

step(red_model)

step(white_model)

step(combined_model)



```


## Anaylsis for Ratings Bands

```{r}
#Filtering for 1-3 Ratings

low_rating <- wine_data %>% filter(quality >= 1 & quality <= 3)
low_rating_red <- red_wine %>% filter(quality >= 1 & quality <= 3)
low_rating_white <- white_wine %>% filter(quality >= 1 & quality <= 3)

high_rating <- wine_data %>% filter(quality >= 8 & quality <= 10)
high_rating_red <- red_wine %>% filter(quality >= 8 & quality <= 10)
high_rating_white <- white_wine %>% filter(quality >= 8 & quality <= 10)

middle_rating <- filtered_wine_data
middle_rating_red <- filtered_red_wine
middle_rating_white <- filtered_white_wine


head(low_rating)

#Comparision combined ratings for red and white

combined_rating_bands <- bind_rows(
  data.frame(type = "low", low_rating),
  data.frame(type = "middle", middle_rating),
  data.frame(type = "high", high_rating)
)

# Arrange the data by type
combined_rating_bands <- combined_rating_bands %>% arrange(type)

# Calculate means for each variable by type
means_by_type <- combined_rating_bands %>%
  group_by(type) %>%
  summarise_all(mean)%>%
  arrange(factor(type, levels = c("low", "middle", "high")))


# Print the means by type
print(means_by_type)



#Comparison combined ratings for red

combined_rating_bands_red <- bind_rows(
  data.frame(type = "low", low_rating_red),
  data.frame(type = "middle", middle_rating_red),
  data.frame(type = "high", high_rating_red)
)

# Arrange the data by type
combined_rating_bands_red <- combined_rating_bands_red %>% arrange(type)

# Calculate means for each variable by type
means_by_type_red <- combined_rating_bands_red %>%
  group_by(type) %>%
  summarise_all(mean)%>%
  arrange(factor(type, levels = c("low", "middle", "high")))


# Print the means by Red type
print(means_by_type_red)



#Comparison combined ratings for white

combined_rating_bands_white <- bind_rows(
  data.frame(type = "low", low_rating_white),
  data.frame(type = "middle", middle_rating_white),
  data.frame(type = "high", high_rating_white)
)

# Arrange the data by type
combined_rating_bands_white <- combined_rating_bands_white %>% arrange(type)

# Calculate means for each variable by type
means_by_type_white <- combined_rating_bands_white %>%
  group_by(type) %>%
  summarise_all(mean)%>%
  arrange(factor(type, levels = c("low", "middle", "high")))


# Print the means by White type
print(means_by_type_white)

```

#Research Questions:
What are the factors that influence the quality rating for wine?  Is there a difference between red and white wine factors?

Furthermore, it appears that the outliers all fall within the 3, 8 and 9 ratings.  What is the difference between those wines, that contributes their place at the top or bottom of the ratings?


#Step Plots – Significant Variables:
Combined – Fixed acidity, volatile acidity, citric acid, residual sugar, free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, & alcohol.

Red – volatile acidity, citric acid, free sulfur dioxide, total sulfur dioxide, pH, sulphates, & alcohol

White – Fixed acidity, volatile acidity, residual sugar, chlorides, free sulfur dioxide, density, pH, sulphates, & alcohol.


#Breakdown by variable

Alcohol:
Alcohol has a positive correlation with the quality rating for both red and white wines.

Sulphates:
When looking at reds and whites together, sulphates do not seem to influence the quality rating.  However, when reds and whites are analyzed separately, it appears that red wines that have a higher sulphate content also have a higher quality rating.

pH: 
pH levels do not have much of effect on quality level.

Density:
The actual difference between values is negligible, however, it appears that as density decreases, the quality rating increases.

Total Sulfur Dioxide and Free Sulfur Dioxide:
For whites, the higher the total sulfur dioxide, the lowered the quality rating.  

Chlorides:
With both, the higher the chlorides, the lower the quality rating.

Residual Sugar:
Residual sugar seems to make a difference with white wines.  The lower the residual sugar, the higher the quality rating.

Citric Acid:
When looking at both together, it seems citric acid has an effect on the quality rating, however, when separating them out, it seems to have a larger effect on reds.   The higher the citric acid the higher the quality rating.

Volatile Acidity 
When looking at both, it appears that the higher the volatile acidity, the lower the quality rating.  When separating them out, that is true for the reds, but not much of an effect on whites.

Fixed Acidity:
When looking at both, it appears that the higher the volatile acidity, the lower the quality rating. When separating them out, that is true for the whites, but does not seem to make a difference for the reds.

#Simplified Conclusion
The factors that influence the quality ratings vary between red and white wine.  Both types share volatile acidity, free sulfur dioxide, pH, sulphates, & alcohol, with alcohol having the strongest positive correlation for both.  However, each type has additional factors that do not influence the other.  For red, it's citric acid, and total sulfur dioxide.  For white, it's residual sugar, chlorides, and density.


